---
title: "BMEG 424 Final Project Report"
author: "Raghav Madhwal (44874626), Alan Khakimov (37431558)"
date: "2026-03-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

The project report revisits and builds upon a differential expression (DE) analysis by RNA-seq to compare a murine erythroleukemia cell line (MEL) to a differentiation resistant derived cell line (MEL-R) [Fernández-Calleja et al., 2017]. The original experiment performed RNA isolation on the two cultured cell lines, prepared standard RNA-seq libraries, and sequenced them using the Illumina GAIIx platform as single-end reads. These reads were deposited in the Gene Expression Omnibus (GEO) database (<http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE83567>) and are utilized in this re-analysis.

In the original workflow, the reads went through quality control (FASTX-Toolkit), were pre-processed to remove adapters, and mapped (TopHat) to a 2012 mouse reference genome (GRCm38, 30-04-2012). Finally, Cufflinks was used to perform a transcriptome expression quantification, and Cuffdiff was used for differential expression analysis. (The pipeline code for the original paper was not included).

Since the original experiment was performed in 2017, both the tools used in their pipeline and reference genome are now outdated. Accurate and splice-aware aligners (such as HISAT2) have superseded TopHat, while Cufflinks/Cuffdiffs has been replaced by improved, more robust, and faster approaches, like Salmon. Furthermore, the mouse reference genome has updated to GCRm39, which could affect gene-level results.

This project aims to reanalyze the RNA-Seq data with these modern methods of analysis. The report aims to build the results of Fernández-Calleja et al., 2017 with tools that are more detailed, accurate, and utilize more robust statistical methods. The reference genome used will also be a more recent version, which may provide better downstream results. By comparing the results from the original workflow to the updated pipeline, this report aims to evaluate the impact of advancements in genome bioinformatics on data interpretation.

## 2. Method and Results

FastQC will be utilized, since it provides better QC depth, with a detailed HTML report for analysis.

fastq will be used for the trimming.

2 alignments methods (HISAT2 and Salmon) will be run for alignment, since they are known to be more accurate compared to TopHat. After HISAT2, the data will be sorted and indexed (through Samtools_sort and Samtools_index). 

Salmon provides expression quantification directly, skipping the need for manual quantification. 

DESeq2 in R will be used instead of cuffdiff for DE analysis.

Description of the methods used, what they do, and why you used them

Analysis is appropriate, well motivated, described, interpreted, and correct Steps are properly validated

### a. Pipeline Creation

### b. Quality Control

Quality control measures are appropriate to the data analyzed, described well, and properly motivated

### c. DESeq2

### d. PCA

### e. Clustering

## Conclusion

Well thought out summary of results, appropriate contrasts with original publication.

## References

Fernández-Calleja, V., Hernández, P., Schvartzman, J. B., Lacoba, M. G. de, & Krimer, D. B. (2017). Differential gene expression analysis by RNA-seq reveals the importance of actin cytoskeletal proteins in erythroleukemia cells. PeerJ, 5, e3432. <https://doi.org/10.7717/peerj.3432>

## Appendix: Plots

```{bash, eval=FALSE}

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
